---
title: "Density Estimation"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
```{r echo=FALSE, warning=FALSE, include=FALSE}

library(microbenchmark)
library(tidyverse)
library(Rcpp)
```

## The ingredients for density estimation

- We have the Epanechnikov Kernel given as
$$K(x) = \frac{3}{4}(1-x^2)1_{[-1,1]}(x)$$
 
- We can then implement the density estimation of $K(x)$ using the following formula
$$\hat{f}_h(x) = \frac{1}{hn}\sum_{j=0}^{n}K(\frac{x-x_j}{h})$$

## Vectorized R implementation

```{r, echo=TRUE}
epan_kern_vec <- function(x, h, m = 512) {
	rg = range(x)
	xx <- seq(rg[1] - 3 * h, rg[2] + 3 * h, length.out = m)
	y <- numeric(m)
	
	const = length(x) * h
	y <- sapply(xx, function(z) sum( (abs((z-x)/h) <= 1) * (3/4 * (1 - ((z - x)/(h))^2)))) 
	list(x = xx, y = y/const)
}
```
- We use ``sapply`` to map all of the gridpoints over into the density estimation
- We use the fact that the boolean test ```abs((z-x)/h) <= 1``` gets coerced into a numeric value

## Testing on infrared data
```{r, echo=FALSE}
infrared <- read.table("./infrared.dat", header = TRUE)
f12 <- infrared$F12
logf12 <- log(f12)
```
```{r, echo=TRUE}
par(mfrow=c(1,2))
d1 <- density(logf12, bw = .2, kernel = "epanechnikov")
d2 <- epan_kern_vec(logf12, h = .2 * sqrt(5))
plot(d1, main="Density estimation infrared data")
lines(d2, col="blue")
range(d1$y - d2$y, na.rm = T)
plot(d1$y - d2$y, type="l", main="Error deviation")
```


## Rccp implementation
```{Rcpp, echo=TRUE}
#include <Rcpp.h>
using namespace Rcpp;
using namespace std;

// [[Rcpp::export]]
NumericVector epan_den(NumericVector x, double h, int m = 512)
{
	int n = x.size();
	
	auto max = max_element(begin(x), end(x));
	auto min = min_element(begin(x), end(x));
	
	NumericVector grid(m);
	double l = abs(*max - *min);
		
	for(int i = 1; i < (m - 1); ++i) {
			grid[i] =  *min + (l / m) * i;
	}
	grid[0] = *min;
	grid[m - 1] = *max; 
	
	NumericVector y(m);
	for(int i = 0; i < m; ++i) {
		for(int j = 0; j < n; ++j) {
			double u = (grid[i] - x[j]) / h;
			if(fabs(u) <= 1.0) {
				y[i] += 3 * (1 - pow(u, 2.0)) / (4);
			}
		}
			y[i] = y[i] / (n * h);
	}
	
	return y;
}
```


## Benchmarking vs. density on infrared data
```{r, echo=TRUE, warning=FALSE, error=FALSE}
sourceCpp('density.cpp')
b <- microbenchmark(
	density(logf12, bw = .2, kernel = "epanechnikov"),
	epan_den(logf12, .2 * sqrt(5)),
	epan_kern_vec(logf12, h = .2 * sqrt(5))
)
autoplot(b)
```

## Benchmarking on simulated data
```{r, echo=TRUE, warning=FALSE, error=FALSE}
x <- rnorm(2^13)
bench_epan_vec <- microbenchmark(
	epan_kern_vec(x[1:32], 0.2 * sqrt(5)),
	epan_kern_vec(x[1:64], 0.2 * sqrt(5)),
	epan_kern_vec(x[1:128], 0.2 * sqrt(5)),
	epan_kern_vec(x[1:264], 0.2 * sqrt(5)),
	epan_kern_vec(x[1:512], 0.2 * sqrt(5)),
	epan_kern_vec(x[1:1024], 0.2 * sqrt(5)),
	epan_kern_vec(x[1:2048], 0.2 * sqrt(5)),
	epan_kern_vec(x[1:4096], 0.2 * sqrt(5)),
	epan_kern_vec(x, 0.2 * sqrt(5)), unit="us"
)

bench_dens <- microbenchmark(
  density(x[1:32], 0.2, kernel = "epanechnikov"),
  density(x[1:64], 0.2, kernel = "epanechnikov"),
  density(x[1:128], 0.2, kernel = "epanechnikov"),
  density(x[1:264], 0.2, kernel = "epanechnikov"),
  density(x[1:512], 0.2, kernel = "epanechnikov"),
  density(x[1:1024], 0.2, kernel = "epanechnikov"),
  density(x[1:2048], 0.2, kernel = "epanechnikov"),
  density(x[1:4096], 0.2, kernel = "epanechnikov"),
  density(x, 0.2), unit = "us"
  )

bench_epan <- microbenchmark(
	epan_den(x[1:32], 0.2 * sqrt(5)),
	epan_den(x[1:64], 0.2 * sqrt(5)),
	epan_den(x[1:128], 0.2 * sqrt(5)),
	epan_den(x[1:264], 0.2 * sqrt(5)),
	epan_den(x[1:512], 0.2 * sqrt(5)),
	epan_den(x[1:1024], 0.2 * sqrt(5)),
	epan_den(x[1:2048], 0.2 * sqrt(5)),
	epan_den(x[1:4096], 0.2 * sqrt(5)),
	epan_den(x, 0.2 * sqrt(5)), unit="us"
)

bench_epan_vec_sum <- summary(bench_epan_vec)
bench_epan_sum <- summary(bench_epan)
bench_dens_sum <- summary(bench_dens)
times <- data.frame(x =  2^(5:13), "density" = bench_dens_sum$uq, "r_vectorized" = bench_epan_vec_sum$uq, 'Rcpp' = bench_epan_sum$uq)
long_times <- times[,-1] %>% gather(key = "density_method", value="value")
long_times$n <- rep(times$x, 3)
```
## Computational time
```{r, echo=TRUE}
long_times %>% ggplot(aes(x = n, color = density_method)) + geom_line(aes(y = log(value)))

```

We see that for small inputs the Rcpp implementation is the fastest.
Yet we still have some sort of polynomial time complexity that density does not have. 

## Further optimization? 
We can implement binning to prevent the $o(nm)$ grid evaluations.. (to be done)

## Bandwidth selection
- Plugin method using emperical variance and guessing using mean zero gaussian density
$$h1_n =  (\frac{8\sqrt{\pi} ||K||_{2}^{2}}{3\sigma_{K}^{4}})^{1/5} \sigma n^{-1/5}$$
We estimate $\sigma$ with the emperical variance. 

## Bandwith calculation on infrared data
First observe 
$$\sigma_{K}^2 = \int z^2 (\frac{3}{4}(1-z^2)) dz = \frac{1}{5}$$
And 
$$||K||_{2}^{2} = \int (\frac{3}{4}(1-z^2))^2 dz = \frac{3}{5}$$
We can now implement the bandwidth estimator
```{r, echo=TRUE}
sigma <- sd(logf12) # log transformed infrared data
n <- length(logf12)
h1n <- function(n) {
	(((8 * sqrt(pi) * (3/5) ) / (3 * (1/5)^2 ))^(1/5)) * n^(-1/5) * sigma
}
h1 <- h1n(n)
h1
h1 / sqrt(5)

```


## Silverman's suggestion for sigma
We substitue the emperical variance with 
$$\tilde{\sigma} = min\{\hat{\sigma}, IQR/1.34\}$$
```{r, echo=TRUE}
sigma_tilde <- min(sd(logf12), IQR(logf12) / 1.34)
h2n <- function(n) {
	(((8 * sqrt(pi) * (3/5) ) / (3 * (1/5)^2 ))^(1/5)) * n^(-1/5) * sigma_tilde
}
h2 <- h2n(n)
h2
h2 / sqrt(5)
```

## Generalizing plugin method wtih a function factory
We can build a generalized bandwidth selector using ``integrate``. 
```{r, echo=TRUE}
bw_select <- function(k) {
	# We need to calculate the l2 squared norm and the variance 
	if(!is.function(k))
		stop('k needs to be a function!')
	sig2 <- integrate(function(x) x^2 * k(x), -Inf, Inf)
	k2 <- integrate(function(x) k(x)^2, -Inf, Inf)
	
	bw <- function(n, sigma_est) {
		numerator <- 8 * sqrt(pi) * k2$value
		denom <- 3 * sig2$value^2
		(numerator / denom)^(1/5) * sigma_est * n^(-1/5)
	}
}

```

## Testing generalized function factory
Note since the function argument k gets integrated on the entire real line we need to specify the kernel limits in the definitino of the kernel itself. 
```{r, echo=TRUE}
epan <- function(x) {
	(abs(x) <= 1) * (3/4) * (1 - x^2) # vectorized because integrate is vectorized aswell
}

## First using emperical variance
bw_epan <- bw_select(epan)

## Test on infrared data
all.equal(bw_epan(n, sd(logf12)), h1)
all.equal(bw_epan(n, sigma_tilde), h2)
```

## Comparison wrt. density function
```{r, echo=FALSE}
d <- density(logf12, kernel = "epanechnikov")
d1 <- density(logf12, bw=h1 / sqrt(5), kernel = "epanechnikov")
d2 <- density(logf12, bw=h2 / sqrt(5), kernel = "epanechnikov")
ds <- data.frame('x' = d$x, 'default' = d$y, 'h1' = d1$y, 'h2' = d2$y)
long_ds <- ds[, -1] %>% gather(key = "bandwidth", value="value")
long_ds$x <- rep(ds$x, 3)

```
```{r, echo=TRUE}
long_ds %>% ggplot(aes(x = x, color = bandwidth)) + geom_line(aes(y = value)) +
		ylab("density") + ggtitle('Bandwidth comparison on infrared data')


```

